Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-02-15 output/paper/_si.qmd
Added class EDCs to SM too.

```{r}
targets::tar_load(
  dplyr::matches("balance_(\\d)$"),
  store = store
)
```

::: {#supptbl-balance-1}
```{r balance-1}
balance_1 |>
  tidylog::rename(Exposure = "exposure") |>
  tidylog::group_by(class) |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  tidylog::ungroup() |>
  gt::gt(groupname_col = "class") |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-balance-2}
```{r balance-2}
balance_2 |>
  tidylog::group_by(class) |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  tidylog::ungroup() |>
  gt::gt(groupname_col = "class") |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-balance-3}
```{r balance-3}
balance_3 |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  gt::gt() |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: glucocorticosteroids; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

```{r}
targets::tar_load(
  dplyr::matches("weights_(\\d)$"),
  store = store
)
```

::: {#supptbl-weights-1}
```{r weights-1}
tidy_res_weighted_fits(
  dat_tbl = weights_1$dat_tbl,
  dat_plt = weights_1$dat_plt,
  sa_var = NULL
)$table
```
**Summary statistics of the estimated balancing weights (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-weights-2}
```{r weights-2}
tidy_res_weighted_fits(
  dat_tbl = weights_2$dat_tbl,
  dat_plt = weights_2$dat_plt,
  sa_var = NULL
)$table
```
**Summary statistics of the estimated balancing weights (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids) (HELIX subcohort; 2013-2016).**
:::

@@

::: {#supptbl-weights-1sa}
```{r weights-1sa}
tidy_res_weighted_fits(
  dat_tbl = weights_1SA$dat_tbl,
  dat_plt = weights_1SA$dat_plt,
  sa_var = "e3_sex"
)$table
```
**Summary statistics of the estimated balancing weights for effect modification (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}; modifier: sex) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-weights-2sa}
```{r weights-2sa}
tidy_res_weighted_fits(
  dat_tbl = weights_2SA$dat_tbl,
  dat_plt = weights_2SA$dat_plt,
  sa_var = "e3_sex"
)$table
```
**Summary statistics of the estimated balancing weights for effect modification (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids; modifier: sex) (HELIX subcohort; 2013-2016).**
:::

>>>>

```{r}
targets::tar_load(
  dplyr::matches("balance_(\\d)$"),
  store = store
)

edcs <- myphd::edcs_information() |>
  dplyr::filter(!chem_id %in% c("dedtp", "dmdtp")) |>
  dplyr::select(chem_id, short_name)
```

::: {#supptbl-balance-1}
```{r balance-1}
balance_1 |>
  dplyr::left_join(edcs, by = c("exposure" = "chem_id")) |>
  dplyr::select(-exposure) |>
  dplyr::rename(Exposure = "short_name") |>
  dplyr::relocate(Exposure) |>
  tidylog::group_by(class) |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  tidylog::ungroup() |>
  gt::gt(groupname_col = "class") |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-balance-2}
```{r balance-2}
balance_2 |>
  dplyr::left_join(edcs, by = c("exposure" = "chem_id")) |>
  dplyr::select(-exposure) |>
  dplyr::rename(Exposure = "short_name") |>
  dplyr::relocate(Exposure) |>
  tidylog::group_by(class) |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  tidylog::ungroup() |>
  gt::gt(groupname_col = "class") |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-balance-3}
```{r balance-3}
balance_3 |>
  dplyr::rename(Exposure = "exposure") |>
  dplyr::arrange(dplyr::desc(Adjusted)) |>
  gt::gt() |>
  gt::fmt_number(n_sigfig = 4) |>
  gt::tab_footnote(
    footnote = "Truncated weights.",
    locations = gt::cells_column_labels(columns = Adjusted)
  ) |>
  gt::opt_footnote_marks(
    marks = "letters"
  ) |>
  gt::tab_style(
    style = gt::cell_text(
      weight = "bold"
    ),
    locations = gt::cells_row_groups()
  )
```
**Effective sample size before and after balancing weights estimation (exposures: glucocorticosteroids; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

```{r}
targets::tar_load(
  dplyr::matches("weights_(\\d)$"),
  store = store
)

edcs <- myphd::edcs_information() |>
    dplyr::filter(!chem_id %in% c("dedtp", "dmdtp")) |>
    dplyr::select(chem_id, short_name, class)
```

::: {#supptbl-weights-1}
```{r weights-1}
tidy_res_weighted_fits(
  dat_tbl = weights_1$dat_tbl,
  dat_plt = weights_1$dat_plt,
  sa_var = NULL
)$table |>
  gt::tab_row_group(
    label = "OP pesticide metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "OP pesticide metabolites", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phenols",
    rows = variable %in% as.character(edcs[edcs$class == "Phenols", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phthalate metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "Phthalate metabolites", ]$short_name)
  ) |>
  gt::row_group_order(
    groups = c("OP pesticide metabolites",
               "Phenols",
               "Phthalate metabolites")
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Summary statistics of the estimated balancing weights (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-weights-2}
```{r weights-2}
tidy_res_weighted_fits(
  dat_tbl = weights_2$dat_tbl,
  dat_plt = weights_2$dat_plt,
  sa_var = NULL
)$table |>
  gt::tab_row_group(
    label = "OP pesticide metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "OP pesticide metabolites", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phenols",
    rows = variable %in% as.character(edcs[edcs$class == "Phenols", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phthalate metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "Phthalate metabolites", ]$short_name)
  ) |>
  gt::row_group_order(
    groups = c("OP pesticide metabolites",
               "Phenols",
               "Phthalate metabolites")
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Summary statistics of the estimated balancing weights (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids) (HELIX subcohort; 2013-2016).**
:::

@@

::: {#supptbl-weights-1sa}
```{r weights-1sa}
tidy_res_weighted_fits(
  dat_tbl = weights_1SA$dat_tbl,
  dat_plt = weights_1SA$dat_plt,
  sa_var = "e3_sex"
)$table |>
  gt::tab_row_group(
    label = "OP pesticide metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "OP pesticide metabolites", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phenols",
    rows = variable %in% as.character(edcs[edcs$class == "Phenols", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phthalate metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "Phthalate metabolites", ]$short_name)
  ) |>
  gt::row_group_order(
    groups = c("OP pesticide metabolites",
               "Phenols",
               "Phthalate metabolites")
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Summary statistics of the estimated balancing weights for effect modification (exposures: {{< acr edcs >}}; outcome: {{< acr hitrtse >}}; modifier: sex) (HELIX subcohort; 2013-2016).**
:::

::: {#supptbl-weights-2sa}
```{r weights-2sa}
tidy_res_weighted_fits(
  dat_tbl = weights_2SA$dat_tbl,
  dat_plt = weights_2SA$dat_plt,
  sa_var = "e3_sex"
)$table |>
  gt::tab_row_group(
    label = "OP pesticide metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "OP pesticide metabolites", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phenols",
    rows = variable %in% as.character(edcs[edcs$class == "Phenols", ]$short_name)
  ) |>
  gt::tab_row_group(
    label = "Phthalate metabolites",
    rows = variable %in% as.character(edcs[edcs$class == "Phthalate metabolites", ]$short_name)
  ) |>
  gt::row_group_order(
    groups = c("OP pesticide metabolites",
               "Phenols",
               "Phthalate metabolites")
  ) |>
  gt::tab_style(
    locations = gt::cells_row_groups(group = dplyr::everything()),
    style = list(gt::cell_text(weight = "bold"))
  )
```
**Summary statistics of the estimated balancing weights for effect modification (exposures: {{< acr edcs >}}; outcomes: glucocorticosteroids; modifier: sex) (HELIX subcohort; 2013-2016).**
:::

### 2024-02-15 output/paper/manuscript.qmd
Added class EDCs to SM too.

Epidemiological evidence suggests that exposure to {{< acr edcs >}} may be linked to perturbation of the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children. We used data from the {{< acr helix >}} cohort including 1,297 children aged 6-11. We used the parametric g-formula and {{< acr mcs >}} to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed positive {{< acr mcs >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with positive {{< acr mcs >}} of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}}. Most of the {{< acr cis first_use=true >}} included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`). The {{< acr opp >}} DETP was strongly negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "hitrtse")`).

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "hitrtse")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "hitrtse")`) and MbZP (`r format_res(heffects1, "mbzp", "hitrtse")`). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-1sa.

@@

@Tbl-hypothesis-3 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "hitrtse")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "hitrtse")`). Furthermore, for all exposures, the {{< acr mcs >}} had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-3sa.

>>>>

Epidemiological evidence suggests that exposure to {{< acr edcs >}} may be linked to perturbation of the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children. We used data from the {{< acr helix >}} cohort including 1,297 children aged 6-11. We used the parametric g-formula and {{< acr mcs >}} to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed positive {{< acr mcs >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with positive {{< acr mcs >}} of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}}. Most of the {{< acr cis first_use=true >}} included the null effect, though. Statistically significant effects were observed for the phenol MEPA (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`). The {{< acr opp >}} DETP was strongly negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "HRT-SE")`).

@@

@Tbl-hypothesis-1and2 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for the phenol OXBE (`r format_res(heffects1, "oxbe", "HRT-SE")`) and the phthalate metabolites MEP (`r format_res(heffects1, "mep", "HRT-SE")`) and MbZP (`r format_res(heffects1, "mbzp", "HRT-SE")`). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-1sa.

@@

@Tbl-hypothesis-3 presents the results of a hypothesis test for the difference between estimates of the {{< acr mcs >}} for females and males, for each exposure. Significant differences were present for cortisone production (`r format_res(heffects3, "cortisone prod.", "HRT-SE")`) and corticosterone production (`r format_res(heffects3, "corticost. prod.", "HRT-SE")`). Furthermore, for all exposures, the {{< acr mcs >}} had opposite sign (positive for males and negative for females). The forest plot of the individual {{< acr mcs >}} is presented in @suppfig-marginal-3sa.
